---
title: "Morphological Principal Components Analysis"
output:
  html_document:
    toc: yes
    toc_depth: 4
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Aim and Setup

### Setup

Load required packages:
```{r setup, echo = TRUE, message = FALSE, warning = FALSE}
#load required packages
library(tidyverse) #data manipulation
library(leiocephalus) #this package
library(ggrepel) #labeling species in plots
library(patchwork) #combining figures
library(here) #easy pathing
```

Set some data partition switches
```{r switches for PCA}

#default: w/ SVL and trait correction
#w/o SVL trait correction
#w/ trait correction but no SVL
#w/o trait correction and SVL

#uses SVL to standardize all other traits
svl.correct <- TRUE
#use SVL in the PCA independent of svl.correct
use.svl <- TRUE
```


Call the raw morphological data and IUCN data
```{r raw data}
data(morph_means, package = "leiocephalus")
data(iucn, package = "leiocephalus")
data(trait_key, package = "leiocephalus")
```

Remove the genus from the names for individual species for the morph dataset before merging them
```{r quick tidying}
#rename the species to drop the genus
morph_means <- morph_means %>%
  mutate(species = str_remove(species, pattern = "Leiocephalus "))
```

Now we should quickly check the relationship between maxsvl (trait 1.1) and svl (trait 1) in `morph_means`. 

```{r check maxsvl vs mean svl}
#check the correlation
cor.test(x = morph_means$`1.1`, y = morph_means$`1`, method = "pearson")

#plot to visualize
ggplot(data = morph_means, aes(x = `1.1`, y = `1`)) +
  geom_point() +
  stat_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  labs(x = "Max SVL", y = "Mean Measured SVL")
```

We are also going to isolate just the traits we are interested in for the PCA:

  - 1     SVL
  - 3     Fore-Hind Limb Distance
  - 6     Pelvis Width
  - 7     Pelvis Height
  - 9.1   Total Tail Length
  - 11    Head Length
  - 12    Head Width
  - 13    Head Height
  - 17    Finger IV Metatarsal Length
  - 20    Upperarm Length
  - 21    Forearm Length
  - 23    Toe IV Metatarsal Length
  - 24    Toe IV Width
  - 26    Thigh Length          
  - 27    Shank Length
  
```{r isolate morph traits}
#vector of traits to keep
keep.morph <- c("1","3","6","7","9.1","11","12","13","17","20","21","23","24","26","27")

#filter accordingly
dat.morph <- morph_means %>%
  dplyr::select(colnames(morph_means)[colnames(morph_means) %in% c("species", keep.morph)]) %>%
  as_tibble(.)
```
  
Now merge the datasets

```{r merge morph and iucn, message=FALSE}
#dat.morph <- right_join(iucn, dat.morph, by = c("species" = "species"))
dat.morph <- right_join(iucn, dat.morph, by = c("species" = "species"))
```

Next, we log-transform our trait data ($log_{10}(trait+1)$) to normalize it and then divide by body size (`SVL`) to correct for body size allometry.

```{r transform and correct for allometry}
if(svl.correct == TRUE && use.svl == TRUE){   #default uses both
#log transform all
ln.dat.morph <- log10(dat.morph[,-1:-7]+1)

#divide all trait values other than SVL by SVL and bind back to log10(SVL+1)
ln.dat.morph <- cbind((ln.dat.morph[1]), (ln.dat.morph[,-1] / ln.dat.morph[,1]))

} else if(svl.correct == TRUE && use.svl == FALSE){ #corrects but does not use svl to plot
#log transform all
ln.dat.morph <- log10(dat.morph[,-1:-7]+1)

#divide all trait values other than SVL by SVL
ln.dat.morph <- ln.dat.morph[,-1] / ln.dat.morph[,1]

} else if(svl.correct == FALSE && use.svl == TRUE){ #does not correct but does uses svl to plot
  #log transform all
ln.dat.morph <- log10(dat.morph[,-1:-7]+1)

} else if(svl.correct == FALSE && use.svl == FALSE){  #does neither
#log transform all
  ln.dat.morph <- log10(dat.morph[,-1:-7]+1)

  #drop SVL
  ln.dat.morph <- ln.dat.morph[,-1]

}

#re-attach the species names
ln.dat.morph <- dat.morph %>%
  dplyr::select(species:range) %>%
  bind_cols(., ln.dat.morph)


knitr::kable(ln.dat.morph)
```

## Analyses & Visualization

### Analyses

Now we can actually conduct the PCA itself. Of note, we use `stats::prcomp(center = TRUE, scale. = TRUE)` to center and scale the data before we make it a nice `tibble`. We still hold onto the actual PCA object but do not add any of the identifying data.

```{r PCA time}
#select just the trait data and conduct PCA
dat.pca <- ln.dat.morph %>%
  dplyr::select(-species:-range) %>%
  prcomp(center = TRUE, scale. = TRUE)

#make tibble version of the PCs
dat_pca_tidy <- dplyr::bind_cols(dplyr::select(ln.dat.morph, species:range), as.data.frame(dat.pca$x))
```

Now we can extract some summary data from the PCA to hold onto for later. We can also add the trait key to the loadings.
```{r pca summary}
#store the variances
pca.var <- summary(dat.pca)

#add loadings and key
pca.loadings <- dat.pca$rotation %>%
  as_tibble() %>%
  cbind(trait = rownames(dat.pca$rotation), .) %>%
  left_join(., trait_key, by = c("trait" = "code")) %>%
  dplyr::select(trait, meaning, everything())
```

### Visualization

Now we can visualize the PCA!

First, let's do that classic screeplot

```{r screeplot}
#make df of the var data
  data.frame(PC = 1:length(pca.var$center),
             var_exp = pca.var$importance[2,]
               ) %>%
  ggplot(aes(x = PC, y = var_exp, group = 1)) +
    geom_point() +
    geom_line() +
    scale_x_continuous(breaks = seq(1, 15, 1)) +
    labs(y = "Proportion of Variance Explained") +
    theme_bw()
```

#### Species PCA Only

```{r viz pca, warning=FALSE}
p_m <- ggplot(data = dat_pca_tidy, mapping = aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = redlist2, size = log10(range)), show.legend = T, stroke = 1.25) +
  geom_text_repel(aes(label=species, color = redlist2), min.segment.length = 0.75, box.padding = 1, max.overlaps = Inf) +
  xlim(c(min(dat_pca_tidy$PC1*1.25),max(dat_pca_tidy$PC1*1.25))) +
  ylim(c(min(dat_pca_tidy$PC2*1.25),max(dat_pca_tidy$PC2*1.25))) +
  scale_shape_manual(values =c(2, 0, 18)) +
  #scale_size_manual(values = c(1.5,1.5,2.5)) +
  scale_size_continuous(range = c(1,5)) +
  labs(
       x = paste0("PC 1 - ", round((pca.var$importance[2,1]*100), 2), " %"),
       y = paste0("PC 2 - ", round((pca.var$importance[2,2]*100), 2), " %")
       ) +
  guides(cex = F, stroke = F, size = F, shape = F) +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()
        )

p_m
```

Now that we have the PCA, we need to add some polygons to help visualize lost area.

#### Species with extinction hulls PCA

We first build the extinction hulls that describe each group of lost species:

  - `past_hull`: no species removed
  - `present_hull`: EX species removed
  - `future_hull`: EX and TH species removed

```{r get hulls}
#get the polygons for the time slices
##PAST
past_hull <- dat_pca_tidy %>%
  mutate(island_bank = str_replace(string = island_bank, pattern = "Bahamas and Cuba", replacement = "Cuba")) %>%
  .[chull(.$PC1, .$PC2),]
levels(past_hull$island_bank) <- levels(dat_pca_tidy$island_bank)

##PRESENT
#RM THE EX SPP
present_hull <- dat_pca_tidy %>%
  mutate(island_bank = str_replace(string = island_bank, pattern = "Bahamas and Cuba", replacement = "Cuba")) %>%
  filter(redlist2 != "EX") %>%
  .[chull(.$PC1, .$PC2),]
levels(present_hull$island_bank) <- levels(dat_pca_tidy$island_bank)

##FUTURE
#RM THE EX AND TH SPP
future_hull <- dat_pca_tidy %>%
  mutate(island_bank = str_replace(string = island_bank, pattern = "Bahamas and Cuba", replacement = "Cuba")) %>%
  filter(redlist2 != "EX" & redlist2 != "TH") %>%
  .[chull(.$PC1, .$PC2),]
levels(future_hull$island_bank) <- levels(dat_pca_tidy$island_bank)

#quickly calculate the lost area
((sp::Polygon(present_hull %>% dplyr::select(PC1,PC2))@area) - (sp::Polygon(future_hull %>% dplyr::select(PC1,PC2))@area)) / (sp::Polygon(present_hull %>% dplyr::select(PC1,PC2))@area)


```

We then can replot the PCA with the different hulls as species are lost:

```{r replot PCA with hulls, message=FALSE, warning=FALSE}
p_m12 <- ggplot(data = dat_pca_tidy %>% filter(redlist2 != "EX"), mapping = aes(x = PC1, y = PC2)) +
    geom_polygon(data = future_hull, aes(x = PC1, y = PC2), color = "grey", fill = rgb(190/255,190/255,190/255,150/255), show.legend = F, alpha = 0.5) +
  geom_polygon(data = present_hull, aes(x = PC1, y = PC2), color = "grey", fill = rgb(190/255,190/255,190/255,100/255), show.legend = F, alpha = 0.3, lty = 3) +
  #geom_polygon(data = past_hull, aes(x = PC1, y = PC2), color = "grey", fill = rgb(190/255,190/255,190/255,50/255), show.legend = F, alpha = 0.1, lty = 2) +
  geom_point(aes(shape = redlist2, size = (range), color = notes), show.legend = T, stroke = 1.25) +
  xlim(c(min(dat_pca_tidy$PC1*1.25),max(dat_pca_tidy$PC1*1.25))) +
  ylim(c(min(dat_pca_tidy$PC2*1.25),max(dat_pca_tidy$PC2*1.25))) +
  #scale_shape_manual(values =c(2, 0, 18)) +
  scale_shape_manual(values =c(0, 18)) +
  #scale_size_manual(values = c(1.5,1.5,2.5)) +
  scale_size_continuous(range = c(1,5)) +
  #scale_color_manual(name = "Exotic Status",values = c("black", "black", "red")) +
  scale_color_manual(name = "Exotic Status",values = c("red", "black")) +
  labs(
       #x = paste0("PC 1 - ", round((pca.var$importance[2,1]*100), 2), " %", "\n large body; large digits; large head; wide pelvis"),
       #y = paste0("PC 2 - ", round((pca.var$importance[2,2]*100), 2), " %", "\nshort tail and limbs; wide and short torso")
       x = paste0("PC 1 - ", round((pca.var$importance[2,1]*100), 2), " %"),
       y = paste0("PC 2 - ", round((pca.var$importance[2,2]*100), 2), " %")
       ) +
  guides(cex = F, stroke = F, size = F, shape = F, color = F) +
  guides(shape=guide_legend(title="Threat Status"), size = F) +
  theme_bw() +
  theme(legend.justification = c(0, 1), 
        legend.position = c(0, 1), 
        legend.background = element_rect(fill = "transparent", colour = "transparent"), 
        legend.key = element_rect(fill = "transparent", colour = "transparent"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()
        )

p_m12
```

```{r time slicing}
#TIME SLICE VERSION
 
#we need to recode the dat_pca_tidy data to be past/present/future
dat_pca_tidy_time <- dat_pca_tidy %>%
  mutate(redlist3 = "PAST") 
 
dat_pca_tidy_time <- bind_rows(dat_pca_tidy_time, #past version
            dat_pca_tidy_time %>% filter(redlist2 != "EX") %>% mutate(redlist3 = "PRESENT"), #present version, we will drop the EX species and then recode the redlist3 for this
            dat_pca_tidy_time %>% filter(redlist2 != "EX" & redlist2 != "TH") %>% mutate(redlist3 = "FUTURE")
            ) %>%
  mutate(redlist3 = factor(redlist3, levels = c("PAST", "PRESENT", "FUTURE")))


p_f_time <- dat_pca_tidy_time %>%
ggplot(data = ., mapping = aes(x = PC1, y = PC2)) +
  geom_point(aes(size = (range), color = notes), show.legend = T, stroke = 1.25, shape = 16) +
  xlim(c(min(dat_pca_tidy$PC1*1.75),max(dat_pca_tidy$PC1*1.75))) +
  ylim(c(min(dat_pca_tidy$PC2*1.75),max(dat_pca_tidy$PC2*1.75))) +
  #scale_shape_manual(values =c(2, 0, 18)) +
  #scale_size_manual(values = c(1.5,1.5,2.5)) +
  scale_size_continuous(range = c(1,5)) +
  scale_color_manual(name = "Exotic Status",values = c("black", "black", "red")) +
  labs(
       x = paste0("PC 1 - ", round((pca.var$importance[2,1]*100), 2), " %"),
       y = paste0("PC 2 - ", round((pca.var$importance[2,2]*100), 2), " %")
       ) +
  guides(cex = F, stroke = F, size = F, shape = F, color = F) +
  #guides(shape=guide_legend(title="Threat Status"), size = F) +
  facet_wrap(~redlist3) +
  theme_bw() +
  theme(legend.justification = c(0, 1), 
        legend.position = c(0, 1), 
        legend.background = element_rect(fill = "transparent", colour = "transparent"), 
        legend.key = element_rect(fill = "transparent", colour = "transparent"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()
        )
```

```{r time slicing with hulls}
#get the hulls
#get the polygons for the time slices
##PAST
past_hull <- dat_pca_tidy_time %>%
  filter(redlist3 == "PAST") %>%
  mutate(island_bank = str_replace(string = island_bank, pattern = "Bahamas and Cuba", replacement = "Cuba")) %>%
  .[chull(.$PC1, .$PC2),]
levels(past_hull$island_bank) <- levels(dat_pca_tidy$island_bank)

##PRESENT
#RM THE EX SPP
present_hull <- dat_pca_tidy_time %>%
  filter(redlist3 == "PRESENT") %>%
  mutate(island_bank = str_replace(string = island_bank, pattern = "Bahamas and Cuba", replacement = "Cuba")) %>%
  .[chull(.$PC1, .$PC2),]
levels(present_hull$island_bank) <- levels(dat_pca_tidy$island_bank)

##FUTURE
#RM THE EX AND TH SPP
future_hull <- dat_pca_tidy_time %>%
  filter(redlist3 == "FUTURE") %>%
  mutate(island_bank = str_replace(string = island_bank, pattern = "Bahamas and Cuba", replacement = "Cuba")) %>%
  .[chull(.$PC1, .$PC2),]
levels(future_hull$island_bank) <- levels(dat_pca_tidy$island_bank)

#combine the hulls
all_hulls <- bind_rows(
  past_hull,
  present_hull,
  future_hull
)

#facetted time plot
p1 <- p_f_time +
  geom_polygon(data = all_hulls, aes(x = PC1, y = PC2), color = "grey", fill = rgb(190/255,190/255,190/255,150/255), show.legend = F, alpha = 0.5) +
    geom_point(aes(size = (range), color = notes), show.legend = T, stroke = 1.25, shape = 16)

p1


#now make a version that is just past and future
p_f_pf <- dat_pca_tidy_time %>%
  filter(redlist3 != "PRESENT") %>%
ggplot(data = ., mapping = aes(x = PC1, y = PC2)) +
  geom_point(aes(size = (range), color = notes), show.legend = T, stroke = 1.25, shape = 16) +
  xlim(c(min(dat_pca_tidy$PC1*1.75),max(dat_pca_tidy$PC1*1.75))) +
  ylim(c(min(dat_pca_tidy$PC2*1.75),max(dat_pca_tidy$PC2*1.75))) +
  #scale_shape_manual(values =c(2, 0, 18)) +
  #scale_size_manual(values = c(1.5,1.5,2.5)) +
  scale_size_continuous(range = c(1,5)) +
  scale_color_manual(name = "Exotic Status",values = c("black", "black", "red")) +
  labs(
       x = paste0("PC 1 - ", round((pca.var$importance[2,1]*100), 2), " %"),
       y = paste0("PC 2 - ", round((pca.var$importance[2,2]*100), 2), " %")
       ) +
  guides(cex = F, stroke = F, size = F, shape = F, color = F) +
  #guides(shape=guide_legend(title="Threat Status"), size = F) +
  facet_wrap(~redlist3) +
  theme_bw() +
  theme(legend.justification = c(0, 1), 
        legend.position = c(0, 1), 
        legend.background = element_rect(fill = "transparent", colour = "transparent"), 
        legend.key = element_rect(fill = "transparent", colour = "transparent"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()
        )

p_f_pf +
  geom_polygon(data = all_hulls %>% filter(redlist3 != "PRESENT"), aes(x = PC1, y = PC2), color = "grey", fill = rgb(190/255,190/255,190/255,150/255), show.legend = F, alpha = 0.5) +
    geom_point(aes(size = (range), color = notes), show.legend = T, stroke = 1.25, shape = 16)

```


#### Species with extinction hulls PCA (PC1 x PC2 and PC1 x PC3)

We now need to recreate hulls for PC1 and PC3 together. This code is identical to that above, except it swaps PC2 for PC3.

```{r pc1 and pc3 hulls}
#get the polygons for the time slices
##PAST
past_hull13 <- dat_pca_tidy %>%
  mutate(island_bank = str_replace(string = island_bank, pattern = "Bahamas and Cuba", replacement = "Cuba")) %>%
  .[chull(.$PC1, .$PC3),]
levels(past_hull13$island_bank) <- levels(dat_pca_tidy$island_bank)

##PRESENT
#RM THE EX SPP
present_hull13 <- dat_pca_tidy %>%
  mutate(island_bank = str_replace(string = island_bank, pattern = "Bahamas and Cuba", replacement = "Cuba")) %>%
  filter(redlist2 != "EX") %>%
  .[chull(.$PC1, .$PC3),]
levels(present_hull13$island_bank) <- levels(dat_pca_tidy$island_bank)

##FUTURE
#RM THE EX AND TH SPP
future_hull13 <- dat_pca_tidy %>%
  mutate(island_bank = str_replace(string = island_bank, pattern = "Bahamas and Cuba", replacement = "Cuba")) %>%
  filter(redlist2 != "EX" & redlist2 != "TH") %>%
  .[chull(.$PC1, .$PC3),]
levels(future_hull13$island_bank) <- levels(dat_pca_tidy$island_bank)
```
Now to create a plot for the other PCs

```{r pca13, message=FALSE, warning=FALSE}
p_m13 <- ggplot(data = dat_pca_tidy, mapping = aes(x = PC1, y = PC3)) +
    geom_polygon(data = future_hull13, aes(x = PC1, y = PC3), color = "grey", fill = rgb(190/255,190/255,190/255,150/255), show.legend = F, alpha = 0.5) +
  geom_polygon(data = present_hull13, aes(x = PC1, y = PC3), color = "grey", fill = rgb(190/255,190/255,190/255,100/255), show.legend = F, alpha = 0.3, lty = 3) +
  geom_polygon(data = past_hull13, aes(x = PC1, y = PC3), color = "grey", fill = rgb(190/255,190/255,190/255,50/255), show.legend = F, alpha = 0.1, lty = 2) +
  geom_point(aes(shape = redlist2, size = (range), color = notes), show.legend = T, stroke = 1.25) +
  xlim(c(min(dat_pca_tidy$PC1*1.25),max(dat_pca_tidy$PC1*1.25))) +
  ylim(c(min(dat_pca_tidy$PC3*1.25),max(dat_pca_tidy$PC3*1.25))) +
  scale_shape_manual(values =c(2, 0, 18)) +
  scale_color_manual(name = "Exotic Status",values = c("black", "black", "red")) +
  #scale_size_manual(values = c(1.5,1.5,2.5)) +
  scale_size_continuous(range = c(1,5)) +
  labs(
       x = paste0("PC 1 - ", round((pca.var$importance[2,1]*100), 2), " %"),
       y = paste0("PC 3 - ", round((pca.var$importance[2,3]*100), 2), " %")
       ) +
  guides(cex = F, stroke = F, size = F, shape = F, color = F) +
  theme_bw() +
  theme(legend.justification = c(0, 1), 
        legend.position = c(0, 1), 
        legend.background = element_rect(fill = "transparent", colour = "transparent"), 
        legend.key = element_rect(fill = "transparent", colour = "transparent"), 
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()
        )
```


We can plot the two PCAs together.

```{r multiPCA}
p_m12 / p_m13
```

Now we can load and try to make the composite figure.

```{r load objects for composite plot, fig.height=5.5, fig.width=7.5}
#load other plots objects
data("extinction_pc1_best")
data("extinction_pc2_best")

#present PCA
pca_present <- dat_pca_tidy_time %>%
    filter(redlist3 != "PAST") %>%
    ggplot(data = ., mapping = aes(x = PC1, y = PC2)) +
    geom_polygon(data = present_hull, aes(x = PC1, y = PC2), color = "grey", fill = rgb(190/255,190/255,190/255,100/255), show.legend = F, alpha = 0.3, lty = 3) +
    geom_point(aes(size = (range), color = notes), show.legend = T, stroke = 1.25, shape = 16) +
    xlim(c(min(dat_pca_tidy$PC1*1.5),max(dat_pca_tidy$PC1*1.5))) +
    ylim(c(min(dat_pca_tidy$PC2*1.5),max(dat_pca_tidy$PC2*1.5))) +
    #scale_shape_manual(values =c(2, 0, 18)) +
    #scale_size_manual(values = c(1.5,1.5,2.5)) +
    scale_size_continuous(range = c(1,5)) +
    #scale_color_manual(name = "Exotic Status",values = c("black", "black", "red")) +
    scale_color_manual(name = "Exotic Status",values = c("red", "black")) +
    labs(
        x = paste0("PC 1 - ", round((pca.var$importance[2,1]*100), 2), " %"),
        y = paste0("PC 2 - ", round((pca.var$importance[2,2]*100), 2), " %")
    ) +
    guides(cex = F, stroke = F, size = F, shape = F, color = F) +
    #guides(shape=guide_legend(title="Threat Status"), size = F) +
    #facet_wrap(~redlist3) +
    theme_bw() +
    ggtitle(label = "A") +
    theme(legend.justification = c(0, 1), 
          legend.position = c(0, 1), 
          legend.background = element_rect(fill = "transparent", colour = "transparent"), 
          legend.key = element_rect(fill = "transparent", colour = "transparent"), 
          panel.grid.minor = element_blank(),
          panel.grid.major = element_blank()
    ) 

#future PCA
pca_future <- dat_pca_tidy_time %>%
    filter(redlist3 == "FUTURE") %>%
    ggplot(data = ., mapping = aes(x = PC1, y = PC2)) +
    geom_polygon(data = future_hull, aes(x = PC1, y = PC2), color = "grey", fill = rgb(190/255,190/255,190/255,150/255), show.legend = F, alpha = 0.5) +
    geom_point(aes(size = (range), color = notes), show.legend = T, stroke = 1.25, shape = 16) +
    xlim(c(min(dat_pca_tidy$PC1*1.5),max(dat_pca_tidy$PC1*1.5))) +
    ylim(c(min(dat_pca_tidy$PC2*1.5),max(dat_pca_tidy$PC2*1.5))) +
    #scale_shape_manual(values =c(2, 0, 18)) +
    #scale_size_manual(values = c(1.5,1.5,2.5)) +
    scale_size_continuous(range = c(1,5)) +
    #scale_color_manual(name = "Exotic Status",values = c("black", "black", "red")) +
    scale_color_manual(name = "Exotic Status",values = c("red", "black")) +
    labs(
        x = paste0("PC 1 - ", round((pca.var$importance[2,1]*100), 2), " %"),
        y = paste0("PC 2 - ", round((pca.var$importance[2,2]*100), 2), " %")
    ) +
    guides(cex = F, stroke = F, size = F, shape = F, color = F) +
    #guides(shape=guide_legend(title="Threat Status"), size = F) +
    #facet_wrap(~redlist3) +
    theme_bw() +
    ggtitle(label = "B") +
    theme(legend.justification = c(0, 1), 
          legend.position = c(0, 1), 
          legend.background = element_rect(fill = "transparent", colour = "transparent"), 
          legend.key = element_rect(fill = "transparent", colour = "transparent"), 
          panel.grid.minor = element_blank(),
          panel.grid.major = element_blank()
    ) 

pca_present + (pca_future + 
  inset_element(p = extinction_pc1_best,
                 left = 0.55,
                   bottom = 0.00, 
                   right = 1.005, 
                   top = 0.35
                ) +
  inset_element(p = extinction_pc2_best,
                   left = 0.00,
                   bottom = 0.75, 
                   right = 0.4550, 
                   top = 1.025
                )
)

ggsave(pca_present + (pca_future + inset_element(p = extinction_pc1_best,
                 left = 0.55,
                   bottom = 0.00, 
                   right = 1.005, 
                   top = 0.35
                ) +
  inset_element(p = extinction_pc2_best,
                   left = 0.00,
                   bottom = 0.75, 
                   right = 0.4550, 
                   top = 1.025
                )), filename = file.path(here(), "figures", "mPCA_extinct_inset.png"),width = 10, height = 8, units = "in", dpi = 300
                #)), filename = file.path(here(), "figures", "mPCA_extinct_inset.png"),width = 1300*(300/72), height = 800*(300/72), units = "px"
  )

#alt equal size panels
#((pca_present + ggtitle("Present")) + (pca_future + ggtitle("Future"))) / ((extinction_pc1_best + ggtitle("PC1 Loss")) + (extinction_pc2_best + ggtitle("PC2 Loss")))

```

## Saving outputs

```{r save pca loadings, eval=FALSE}
#need to se the way the data are saved here based on the switches added
if(svl.correct == TRUE && use.svl == TRUE){
  #default uses both
save(dat_pca_tidy, file = file.path(here::here(), "data", "dat_pca_tidy.rda"))
} else if(svl.correct == TRUE && use.svl == FALSE){
  #rename for saving
  dat_pca_tidy_nosvl <- dat_pca_tidy
  #corrects but does not use svl to plot
save(dat_pca_tidy_nosvl, file = file.path(here::here(), "data", "dat_pca_tidy_nosvl.rda"))
} else if(svl.correct == FALSE && use.svl == TRUE){
    #rename for saving
  dat_pca_tidy_norescale <- dat_pca_tidy
  #does not correct but does uses svl to plot
save(dat_pca_tidy_norescale, file = file.path(here::here(), "data", "dat_pca_tidy_norescale.rda"))
} else if(svl.correct == FALSE && use.svl == FALSE){
      #rename for saving
  dat_pca_tidy_nosvl_norescale <- dat_pca_tidy
  #does neither
save(dat_pca_tidy_nosvl_norescale, file = file.path(here::here(), "data", "dat_pca_tidy_nosvl_norescale.rda"))
}
```
